### 2018-05-29 Paper/jaccardbaetica.tex
add retoques

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of those factors have an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

@@

Roman province such as Baetica (currently Andalusia, south Spain) became important production centers of olive oil in the Roman Age. Olive oil was an important product during the Roman Empire used in almost all the aspect of daily life as cooking, hygiene or lightning \citep{mattingly_d.j._oil_1988}.

The high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the rivers supplying from the riverine connectivity to the Mediterranean and Atlantic \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps and \textit{tituli picti} and inked in graffities with different information but there is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. Stamps are the most studied in this type of amphorae. There is evidence that stamps were used for almost three centuries. (economía oleica betica). Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria (CITAR). Stamps was mostly formed for a code of three letters. There letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
The meaning of the amphora stamps is not clear. On the one hand, stamps are identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could be the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
However, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow some distribution pattern? Did stamps share the same workshop? Neither written sources have been found that it could explain the economic role of Baetica province in the Roman organization. On the other hand, archaeological evidence show a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

@@

From the database, we collected the site where it was found and the stamp code. We also created a new column with the area where stamps were found (decir berni). Those areas, known as \textit{conventus}, were administrative centers for territorial organization in the Roman Empire (CITAR). Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalis} (currently Seville), \textit{Corduba} (currently C\'ordoba) and \textit{Astigi} (currently Écija, Sevilla) \citep{rodriguez_economioleicola_1977,chic_datos_2001,berni_millet_epigrafianforica_2008} .

The distribution of amphora stamps in different conventus can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly workshops show a homogeneity on frequency of stamps except La Catria and Arva (both show 29 different stamps) that show a big amount of different stamps. According to some researchers, those workshops became in one of the most important centres of amphora production (CITAR) although it could have been intensely prospected. \citep{arva_1997}

@@

Considering our dataset as non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index can be expressed considering 0 as total presence of similarity of stamps and 1 a totally dissimilarity between stamps. In our case, it will be calculated the number of times that one stamp appear in a amphora workshop. This method allows to bear in mind the similar number of times for each repeated stamp per workshop. 
%quizas ampliar con el método de xavielin

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. This dendrogram suggests that amphora workshops used different stamps for their production system. No groups of similar stamps were found in the cluster. On the other hand, closer workshops show a similarity on the stamps while most of them seem to display different stamps roles. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, as for instance Picachos, Cerro de los Pesebres and El Castillejo.

@@

In this study, we aimed to analyse whether amphoric stamps could play an important role in the organization of the workshops along rivers. For that, dissimilarity index was used to detect differences.

In our analysis, no strong relation between stamps and geographic distance were detected in amphora workshops. Indeed, the analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions. Consequently the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in closer workshops and not in other areas. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

@@

%los que aparecen cerca son porque están cerca o puede haber confusion entre talleres
%By contrast, we have located different stamps in a same amphora workshop.

@@

The research was funded by European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon. 
All data has been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}.

>>>>

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors have an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

@@

Roman province such as Baetica (currently Andalusia, south Spain) became important production centers of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, the high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in graffities with different information. However, graffities and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria. Stamps was mostly formed for a code of three letters. These letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

INCLUIR AQUÍ QUE LAS STAMPS VAN CAMBIANDO DE FECHA Y SIGLOS

%Many hypothesis involved the meaning of stamps. 
 There is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. On the one hand, stamps are identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could be the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow some distribution pattern? Did stamps share the same workshop? Neither written sources have been found that it could explain the economic role of Baetica province in the Roman organization. On the other hand, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

@@

From the database, we collected the site where it was found and the stamp code. We also created a new column with the area where stamps were found (decir berni). Those areas, known as \textit{conventus}, were administrative centers for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalis} (currently Seville), \textit{Corduba} (currently C\'ordoba) and \textit{Astigi} (currently Écija, Sevilla) \citep{rodriguez_economioleicola_1977,chic_datos_2001,berni_millet_epigrafianforica_2008} .

The distribution of amphora stamps in different conventus can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly workshops show a homogeneity on frequency of stamps except La Catria and Arva (both show 29 different stamps) that show a big amount of different stamps. According to some researchers, those workshops became in one of the most important centres of amphora production although it could have been more intensely prospected then others. \citep{arva_1997}

@@

Considering our dataset as non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index can be expressed considering 0 as total presence of similarity of stamps and 1 a totally dissimilarity between stamps. In our case, it will be calculated the number of times that one stamp appear in a amphora workshop. This method allows to bear in mind the similar number of times for each repeated stamp per workshop. If two workshops have similar stamp codes then the probability would be 0 while the results would be 1 when stamp codes are totally different.

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. This dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, no groups of similar stamps were found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, as for instance Picachos, Cerro de los Pesebres and El Castillejo.

@@

In this study, we aimed to analyse whether amphoric stamps could play an important role in the organization of the workshops along rivers. For that, dissimilarity index was used to detect differences among workshops and stamps.

In our analysis, no strong relation between stamps and geographic distance were detected in amphora workshops. Indeed, the analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions when nearby workshops share the same amphoric stamps. Consequently the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in the same conventus area. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

@@

@@

The research was funded by European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon and Nachete. 
All data has been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}.
